A Simple, Useful, and Only Slightly Wrong Model of Virtual Memory

### 2010-08-23 article.tex
Fixed lines

\documentclass{acmtrans2m}
\usepackage{url}
\usepackage{salgorithm}
\usepackage{graphicx} \graphicspath{{sorts/}}
\usepackage{fancyvrb}

%\acmVolume{V} 
%\acmNumber{N} 
%\acmYear{YY} 
%\acmMonth{M} 
\markboth{Peter Boothe}{A Simple, Useful Model of Virtual Memory} 
\title{A Simple, Useful, and Only Slightly Wrong Model of Virtual Memory}
\author{PETER BOOTHE\\
Manhattan College
}
\begin{abstract} 
We define a simple model of virtual memory in an effort to predict the number of page faults that an algorithm will occur over the course of its lifetime.  We then use that model to analyze how many page faults we would expect to see with two of the best-known and most well-studied sorting algorithms.  Then, we run multiple experiments in an effort to show that the predictions of our model match up with reality.

\end{abstract}

@@

\begin{document} 
\begin{bottomstuff} 
This research was supported by a summer grant from Manhattan College.\\
Author's Address: \url{peter.boothe@manhattan.edu}, or P. Boothe, Manhattan College Math and Computer Science Dept. RLC 201, 3840 Corlear Ave, Bronx, NY 10463\\
\end{bottomstuff} 
\maketitle

@@

In the worst case, every memory access will have to go to the hard drive.
This, however, is not an informative worst-case analysis.  The worst-case page
fault predictions do not reflect the real-world behavior of virtual memory ---
they reflect a situation in which every cache access is a cache miss! Our
model predicts that all non-adjacent memory accesses have a uniform non-zero probability of causing a page fault.

@@

Before we continue, however, we should address one obvious fault of our model --- it is not completely correct!  In particular, if we wrote a program that traversed an array that was twice the size of available memory, our model predicts that we will see at most 1 page fault, when in actuality there will be at least $\frac{n}{2}$.  This objection is correct, as far as it goes, and means that our model can be ``gamed'' and should only serve as a lower-bound measure of page faults if exactness is required.  However, the extreme simplicity of our model has a lot of appeal, so we will conduct our experiments and see if the predictions of our (slightly wrong) model are useful and accurate.

@@

\subsection{\heapsort\ Page Fault Analysis}
Our analysis of \heapsort\ is the easiest.  The \heapsort\ algorithm, which we
take from \citeN{clrs}, is defined, with all of its
supporting functions, in Figure~\ref{fig:hsort}.  The implementation we tested
is from the BSD standard library, which is available as {\tt libbsd} on Linux.

@@

We reconcile the opposing forces of our desire for speed and our desire to use
swap by running our tests in a virtual machine via the Qemu emulator.  The
machine running the emulator had 4 gigabytes of memory.  Of those 4 gigabytes,
half was allocated as a RAM disk.  The swap partition of the emulated machine
was a file stored on that RAM disk.  This meant that the page faults of the
emulated machine would still be recorded, but that the emulated machine would
not suffer a performance hit, as the emulated swap was backed by a (very fast)
RAM disk!  This technique is recommended for anyone who would like to count
page faults but does not want to waste a lot of time.

@@

\begin{figure}
\input{sorts/ratios.tex}
\caption{Number major page faults for \heapsort\ divided by the number of major page faults for \mergesort\ at a variety of memory sizes.  The fact that this ratio grows over time indicates the \heapsort\ has asymptotically more page faults than \mergesort, which is exactly what is predicted by our model!}
\end{figure}

@@

\begin{received} 
Received Month Year; revised Month Year; accepted Month Year 
\end{received}
\end{document}

>>>>

\documentclass{acmtrans2m}

\def\endbottomstuff{\end{figure}\gdef\permission{}}

\def\runningfoot{\def\@runningfoot{}}
\def\firstfoot{\def\@firstfoot{}}

\usepackage{url}
\usepackage{salgorithm}
\usepackage{graphicx} \graphicspath{{sorts/}}
\usepackage{fancyvrb}

%\acmVolume{V} 
%\acmNumber{N} 
%\acmYear{YY} 
%\acmMonth{M} 
\markboth{Peter Boothe}{A Simple, Useful Model of Virtual Memory} 
\title{A Simple, Useful, and Only Slightly Wrong Model of Virtual Memory}
\author{PETER BOOTHE\\
Manhattan College
}
\begin{abstract} 
We define a simple model of virtual memory in an effort to predict the number
of page faults that an algorithm will occur over the course of its lifetime.
We then use that model to analyze how many page faults we would expect to see
with two of the best-known and most well-studied sorting algorithms.  Then, we
run multiple experiments in an effort to show that the predictions of our model
match up with reality.
\end{abstract}

@@

\begin{document} 
\begin{bottomstuff} 
This research was supported by a summer grant from Manhattan College.\\
Author's Address: \url{peter.boothe@manhattan.edu}, or P. Boothe, Manhattan College Math and Computer Science Dept. RLC 201, 3840 Corlear Ave, Bronx, NY 10463
\end{bottomstuff} 
\maketitle

@@

In the worst case, every memory access will have to go to the hard drive.
This, however, is not an informative worst-case analysis.  The worst-case page
fault predictions do not reflect the real-world behavior of virtual memory ---
they reflect a situation in which every cache access is a cache miss. Our model
predicts that all non-adjacent memory accesses have a uniform non-zero
probability of causing a page fault.

@@

Before we continue, however, we should address one obvious fault of our model
--- it is not completely correct!  In particular, if we wrote a program that
traversed an array that was twice the size of available memory, our model
predicts that we will see at most 1 page fault, when in actuality there will be
at least $\frac{n}{2}$.  This objection is correct, as far as it goes, and
means that our model can be ``gamed'' and should only serve as a lower-bound
measure of page faults if exactness is required.  However, the extreme
simplicity of our model has a lot of appeal, so we will conduct our experiments
and see if the predictions of our (slightly wrong) model are nonetheless useful
and accurate.

@@

\subsection{\heapsort\ Page Fault Analysis}

Our analysis of \heapsort\ is the easiest.  The \heapsort\ algorithm, which we
take from \citeN{clrs}, is defined, with all of its supporting functions, in
Figure~\ref{fig:hsort}.  The implementation we tested is from the BSD standard
library, which is available as {\tt libbsd} on Linux.

@@

We reconcile the opposing forces of our desire for speed and our desire to use
swap by running our tests in a virtual machine via the Qemu emulator.  The
machine running the emulator had 4 gigabytes of memory.  Of those 4 gigabytes,
half was allocated as a RAM disk.  The swap partition of the emulated machine
was a file stored on that RAM disk.  This meant that the page faults of the
emulated machine would still be recorded, but that the emulated machine would
not suffer a performance hit, as the emulated swap was backed by a (very fast)
RAM disk.  This technique is recommended for anyone who would like to count
page faults but does not want to waste a lot of time.

@@

\begin{figure}
\input{sorts/ratios.tex}
\caption{Number major page faults for \heapsort\ divided by the number of major page faults for \mergesort\ at a variety of memory sizes.  The fact that this ratio grows over time indicates the \heapsort\ has asymptotically more page faults than \mergesort, which is exactly what is predicted by our model.}
\end{figure}

@@

\begin{received} 
Submission to ALENEX 2011
\end{received}
\end{document}
